---
title: Exploratory Analysis of Time Series Data with Tidyverts
author: Dennis Sobolewski
date: '2020-02-10'
slug: exploratory-analysis-of-time-series-data-with-tidyverts
categories:
  - R
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2020-02-10T15:48:31-06:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include=FALSE}
library(tidyverse)
library(fpp3)


knitr::opts_chunk$set(fig.retina = 3, warning = FALSE, message = FALSE, echo = FALSE)


tourism <- read_csv(here::here("static","data","tourism.csv"))

```

## Into to Tidyverts

Last month, I attended rstudio::conf 2020 and took [Rob J Hyndman's](https://robjhyndman.com/) awesome Tidy Time Series and Forecasting in R workshop. Professor Hyndman highlighted functionality within the [Tidyverts](https://tidyverts.org/) packages for exploring and extracting features from time series datasets.Tidyverts is currently comprised of three main packages and works within the Tidyverse framework (i.e. piping and dplyr functions).

* __Tsibble-__ Makes data wrangling and formatting of time series data easier. Formats time series data into a "tsibble" R object so other packages within Tidyverts know how to handle it.
* __Feasts-__ Used for extracting features from time series data. Has many useful functions for extracting statistics from a time series that can be used for exploratory analysis, model checking, and comparison.
* __Fable-__ Simplifies the creation for forecasting models for time series data.

I will review and expand on the exploratory data analysis techniques, primarily using tsibble and feasts, learned at rstudio::conf in this post. The workshop changed my perspective on how time series data can be explored, from a simplistic format with limited options (pretty much just line graphs) to a still growing discipline with room for creative solutions. Tidyverts lets you squeeze more insight from a simple time series than I thought possible.

## Creating a Tsibble Object

To begin using the Tidyvert packages, you first need to convert your data into a tsibble object. I will demonstrate this using a dataset showing quarterly Australian tourism totals by State, Region, and purpose that can be found [here](http://robjhyndman.com/data/tourism.xlsx). 

```{r data_check, include=TRUE, echo = TRUE}

head(tourism)

```

The tibble has a Quarter column containing dates that will represent our time variable. Before creating a tsibble you need to make sure the time variable is converted to the correct interval. If you set the index equal to the Quarter variable before transforming it, our tsibble functions will assume the desired interval length is one day with three months of data is missing between observations. The tsibble package makes it easy to change this column from a date to a "qtr" data type using the yearquarter() function. Each variable that uniquely determines a time series we wish to measure should be specified using the "key" parameter. 

```{r tsibble_obj, include=TRUE, echo = TRUE}

tourism_ts <- tourism %>%
  mutate(Quarter = tsibble::yearquarter(Quarter)) %>%
  tsibble::as_tsibble(index = Quarter, key = c(Region, State, Purpose)) 

head(tourism_ts)

```

Inspecting the new object displays the key variables and shows the updated data type of the Quarter column. The functions within all of the Tidyverts packages are built to work with tsibble objects based on the index and key specification. Most functions will iterate over each key combination present in the dataset by default. What you would typically use an apply or purrr function to achieve just happens automatically. This makes it simple to visualize and run calculations on many different time series at once. It's almost disconcerting to see how simple the Tidyverts framework makes time series analysis, especially if you ever worked with creating and working with ts objects in the past. It feels like you're cheating!

## Visualizing Time Series

Tsibble objects have their own preset ggplot when using the autoplot() function. Autoplot will create a time series line graph for each unique key combination in our tsibble object by default. The interval being used is always specified at the bottom. 

```{r autoplot, include=TRUE, echo = TRUE}

autoplot(tourism_ts, Trips) + 
  guides(color = F) 


```

Since these objects interact nicely with the rest of the tidyverse, dplyr functions can be used to filter, group, and summarise tsibbles to explore the data further. Also to note, autoplot() is just a pre-defined ggplot output based on the data being input and all normal ggplot syntax can be used after to change the look of your plots. 

```{r autoplot_dplyr, include=TRUE, echo = TRUE}

holidays_ts <- tourism_ts %>%
  filter(Purpose == "Holiday") %>%
  group_by(State) %>%
  summarise(Trips = sum(Trips))

holidays_ts %>%
  autoplot(Trips) +
  labs(title = "Australian Holiday Travel", col = "") +
  theme(legend.position="bottom") +
  guides(col=guide_legend(ncol=4)) 
  
```

The Feasts package is full of custom ggplot visualization functions that play nicely with tsibble objects and are super useful for data exploration.

### Seasonal Plots

The gg_season() function will create a plot that chops our time series data into even periods and plots them on top of each other. This lets us better see trends in our data over a given period. In the below example I took our holiday_ts object, filtered it to the state of Victoria, and created a seasonal plot. It is clear a seasonal trend exists with Q3 typically being the lowest travel quarter when we look at our data from this new perspective.

```{r gg_season, include=TRUE, echo = TRUE}

holidays_ts %>%
  filter(State == "Victoria") %>%
  feasts::gg_season(Trips, labels = "right")

```

If we don't filter the data first, the gg_season() function will automatically facet our plot by every key combination. This lets us visually compare multiple time series when exploring our data. We can see the states with trends similar to Victoria but also those with an opposite trend and high Q3 holiday travel volume. 

```{r gg_season_facet, include=TRUE, echo = TRUE, fig.height=10}

holidays_ts %>%
  feasts::gg_season(Trips)
  
```

If the tsibble has a smaller interval, the period being shown on the bottom of this graph can be adjusted. Below I used the tsibbledata::vic_elec dataset that has Victoria household electricity usage down to the half hour. By setting period = "day" in gg_season(), an hourly line is drawn for each day in the dataset allowing us to see typical peak usage times. 

```{r gg_season_hourly, include=TRUE, echo = TRUE}

tsibbledata::vic_elec %>%
  feasts::gg_season(Demand, period = "day")

```

### Subseries Plots

Another useful plotting function is feasts::gg_subseries(). This will facet the entire series by a smaller period to allow you to see trends within those subsets. In the below example we can see that holiday trips to Victoria are clearly increasing at a much greater rate in the 1st and 4th quarters compared to the 2nd and 3rd.

```{r gg_subseries, include=TRUE, echo = TRUE}

holidays_ts %>%
  filter(State == "Victoria") %>%
  feasts::gg_subseries(Trips)

```

Once again if you do not filter first the graph is further faceted by the key combinations in the dataset to allow for more detailed visual comparison. 

```{r gg_subseries_facet, include=TRUE, echo = TRUE, fig.height=10}

holidays_ts %>%
  feasts::gg_subseries(Trips)
  
```

### Calendar Plots

Earo Wang's [sugrrants](https://github.com/earowang/sugrrants) package is not a part of Tidyverts but it helps us create calendar plots that are just too pretty to be left out of this post. Since this is not part of tidyverts it actually is easier to manipulate the data as a tibble. There are two flavors of this plot shown below. The second is particularly useful if you are trying to show more than a few months worth of data in a calendar format.

```{r calendar_hourly1, include=TRUE, echo = TRUE, fig.height=10}

tsibbledata::vic_elec %>%
  as_tibble %>%
  mutate(Hour = lubridate::hour(Time)) %>%
  filter(lubridate::year(Date) == 2013,
         lubridate::month(Date) %in% c(4,5)) %>%
  group_by(Date,Hour) %>%
  summarise(Demand = sum(Demand)) %>%
  mutate(Weekend = if_else(lubridate::wday(Date)  %in% c(1,7), "Weekend", "Weekday")) %>%
  ggplot(aes(x = Hour, y = Demand, col = Weekend)) +
  geom_line() +
  sugrrants::facet_calendar(~ Date, ncol = 1) +
  theme_bw() +
  theme(legend.position = "bottom")

```

```{r calendar_hourly2, include=TRUE, echo = TRUE, fig.height=10}

calendar <- tsibbledata::vic_elec %>%
  filter(lubridate::year(Date) == 2013) %>%
  mutate(Hour = lubridate::hour(Time)) %>%
  group_by(Date,Hour) %>%
  summarise(Demand = sum(Demand)) %>%
  mutate(Weekend = if_else(lubridate::wday(Date) %in% c(1,7), "Weekend", "Weekday")) %>%
  sugrrants::frame_calendar(
    x = Hour, 
    y = Demand, 
    date = Date, 
    nrow = 4
  ) %>%
  ggplot(aes(x = .Hour, y = .Demand, group = Date, col = Weekend)) +
  geom_line() +
  theme(legend.position = "bottom")

sugrrants::prettify(calendar, size = 3, label.padding = unit(0.15, "lines"))


```

### ACF Plots

Auto-corrections of lagged values for a time series can provide valuable insight in to seasonal or cyclical trends present in the data. The feasts package a an ACF() function that will provide these auto-correlation values for a tsibble object. 

```{r acf1, include=TRUE, echo = TRUE}

holidays_ts %>%
  filter(State == "Victoria") %>%
  feasts::ACF(Trips, lag_max = 12) 

```

Feeding these values into the autoplot() function will create an ACF plot that includes a confidence band to help determine significant autocorrelations. Time series with auto-correlations outside of the confidnce bands at constant intervals indicate seasonality. Creating ACF plots on the residuals left over from a forecast can help determing if there are remaining patterns in the data that your model is not accounting for. 

```{r acf2, include=TRUE, echo = TRUE, fig.height=10}

holidays_ts %>%
  feasts::ACF(Trips) %>%
  autoplot()


```

## Feature Extraction

The Tidyverts package includes functions that make it easy to extract many different features from atsibble object. The fabletools::features() function allows you to simply choose the feature, or set of feastures, that you wish to extract. Feasts contains pre-determined sets of features taht can be found with ?fabletools::features_by_pkg(). The below example uses the feasts::feat_stl feature set which results in nine features being extracted that summarise each time series.

```{r stl_feats, include=TRUE, echo = TRUE}

tourism_stl_feats <- tourism_ts %>%
  fabletools::features(Trips, features = feasts::feat_stl) 

head(tourism_stl_feats)

```

This is powerful for inspecting your time series in creative ways. The below plot looks at the seasonal and overall trend strength of each time series faceted by state and colored by travel purpose. This lets you see patterns across states not possible with traditional time series plots. 

```{r stl_feats_plot, include=TRUE, echo = TRUE}

tourism_stl_feats %>%
  ggplot(aes(x = trend_strength, y = seasonal_strength_year, col = Purpose)) +
  geom_point() +
  facet_wrap(~State)

```

You can further use these features to answer specific questions of your time series data. For example, below filters our feature set to the time series with the strongest seasonal pattern. Holiday trips in the snowy mountains of New South Wales have the most seasonality in our dataset. 

```{r stl_feats_seasonal, include=TRUE, echo = TRUE}

most_seasonal <- tourism_stl_feats %>%
  filter(seasonal_strength_year == max(seasonal_strength_year))

tourism_ts %>%
  inner_join(most_seasonal, by = c("State", "Region", "Purpose")) %>%
  ggplot(aes(x = Quarter, y = Trips)) + geom_line() +
  labs(title = "Most Seasonal Series") +
  facet_grid(vars(State, Region, Purpose))

```

The same can be done for overall trend. Business trips in Western Australia have the strongest positive overall trend in our dataset. 

```{r stl_trends_seasonal, include=TRUE, echo = TRUE}

most_trended <- tourism_ts %>%
  fabletools::features(Trips, feasts::feat_stl) %>%
  filter(trend_strength == max(trend_strength))

tourism_ts %>%
  inner_join(most_trended, by = c("State", "Region", "Purpose")) %>%
  ggplot(aes(x = Quarter, y = Trips)) + geom_line() +
  labs(title = "Most Trended Series") +
  facet_grid(vars(State, Region, Purpose))

```

### Hierarchical Clustering and PCA Analysis

The ability to extract many different features from multiple time series at once opens up many possibilities. These features can be used in PCA or clustering analyses to provide different insights from our data. 

The fabletools::features allows you to pull every feature available in the feasts pacakge by using the fabletools::feature_set function. This will result in 44 features being calculated for each time series. I'm not going to lie- I don't know what all of these features represent or how they are calculated, but I'm not going to let that stop me from using them for this example.

```{r tourism_features, include=TRUE, echo = TRUE}

tourism_features <- tourism_ts %>%
  fabletools::features(Trips, fabletools::feature_set(pkgs = "feasts"))

head(tourism_features)

```

Performing a silhouette analysis reveals an optimal K value of 3 for clustering. A dendrogram of our hierarchical clustering results using Euclidean distance measurements is below. 

```{r tourism_features_cluster, include=TRUE, echo = FALSE}

features_matrix <- tourism_features %>%
  select(-State, -Region, -Purpose) %>%
  data.matrix() %>%
  scale() 

row.names(features_matrix) <- paste0(tourism_features$State,"/",tourism_features$Region,"/",tourism_features$Purpose)


##Silhouette Analysis
sil_width <- map_dbl(2:24, function(k){
  model <- cluster::pam(features_matrix, k = k)
  model$silinfo$avg.width
  
})

data.frame(k = 2:24, sil_width = sil_width) %>%
  ggplot(aes(x = k, y = sil_width)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = 2:24) +
  labs(title = "Silhouette Anaylsis for Optimal K")

## Hierarchical clustering

features_hc <- hclust(dist(features_matrix, method = "euclidean"), method = "complete")

dendextend::color_branches(as.dendrogram(features_hc), k = 3) -> features_dend

plot(features_dend, leaflab = "none")
```

### Cluster 3

The red cluster contains only 2 time series from our original data set that were different enough to necessiate their own cluster. Plotting these reveals they have irregular looking seasonality and strong positive trends overall. 

```{r tourism_cluster_3, include=TRUE, echo = FALSE}

set.seed(1988)
tourism_features %>%
  mutate(cluster = cutree(features_hc, k = 3)) -> features_clust

tourism_ts %>%
  inner_join(features_clust) %>%
  filter(cluster == 3) -> cluster_3

cluster_3 %>%
  mutate(Series = paste0(State,"/",Region,"/",Purpose)) %>%
  ggplot(aes(x = Quarter, y = Trips)) +
  geom_line() +
  labs(title = "Outlying time series in HC space (Cluster 3)") +
  facet_wrap(~Series,ncol = 1, scales = "free")

cluster_3 %>%
  feasts::gg_season()

cluster_3 %>%
  feasts::gg_subseries()

```

### Clusters 1 and 2

The other two larger clusters contain more time series. Lets look at how they most differ from each other. The below shows the top features with the largest difference between the two clusters with spikiness being the number one differentiating feature. 

```{r tourism_cluster_others, include=TRUE, echo = FALSE}

cluster_summary <- features_clust %>%
  select(-State,-Region,-Purpose) %>%
  group_by(cluster) %>%
  summarise_all(mean) %>%
  pivot_longer(cols = -cluster) %>%
  pivot_wider(names_from = cluster,
              values_from = value) 

scaled <- cluster_summary %>%
  select(`1`,`2`) %>%
  data.matrix() %>%
  scale()

cluster_summary %>%
  select(-`3`) %>%
  cbind(scaled_abs_diff = abs(as.numeric(scaled[,2])-as.numeric(scaled[,1]))) %>%
  arrange(desc(scaled_abs_diff)) %>%
  head()

```

```{r tourism_cluster_others, include=TRUE, echo = FALSE}

tourism_ts %>%
  inner_join(features_clust) %>%
  autoplot(Trips) +
  guides(color = F) +
  facet_wrap(~cluster, ncol = 1) +
  labs(title = "Australia Trips Faceted by Cluster")

tourism_ts %>%
  inner_join(features_clust) %>%
  group_by(cluster) %>%
  tsibble::index_by(Quarter) %>%
  summarise(Trips = mean(Trips)) %>%
  autoplot(Trips) +
  labs(title = "Average Trip Total by Cluster")

```


```{r tourism_features_pca, include=TRUE, echo = TRUE}

## PCA Analysis


features_pcs <- tourism_features %>%
  select(-State, -Region, -Purpose) %>%
  prcomp(scale = TRUE) %>%
  broom::augment(tourism_features)


features_pcs %>% 
  ggplot(aes(x=.fittedPC1, y=.fittedPC2, col = Purpose)) +
  geom_point(size = 1.5) + 
  theme(aspect.ratio=1)


features_pcs %>% 
  ggplot(aes(x=.fittedPC1, y=.fittedPC2, col = Purpose)) +
  geom_point(aes(size = .fittedPC1 > 11)) + 
  scale_size_manual(values =c(1.5, 4), guide = F)+
  theme(aspect.ratio=1)


outliers <- features_pcs %>%
  filter(.fittedPC1 > 11)

outliers %>%
  left_join(tourism, by = c("State", "Region", "Purpose")) %>%
  mutate(Series = paste0(State,"/",Region,"/",Purpose)) %>%
  ggplot(aes(x = Quarter, y = Trips)) + 
  geom_line() +
  labs(title = "Outlying time series in PC space") +
  facet_wrap(~Series, ncol = 1, scales = "free")

```

```{r tourism_features_pca_clust, include=TRUE, echo = TRUE}

features_pcs %>% 
  inner_join(features_clust) %>%
  ggplot(aes(x=.fittedPC1, y=.fittedPC2, col = factor(cluster))) +
  geom_point() + 
  labs(color = "Cluster") +
  theme(aspect.ratio=1)


features_pcs %>% 
  inner_join(features_clust) %>%
  ggplot(aes(x=.fittedPC1, y=.fittedPC2, col = Purpose)) +
  geom_point() + 
  scale_size_manual(values =c(1.5, 4), guide = F)+
  facet_wrap(~cluster) +
  labs(title = "PC Space Plot Faceted by Cluster") +
  theme(legend.position="bottom", aspect.ratio=1) 


```